---
title: "TidyVerse Create and Extend Vignette"
output:
  pdf_document: 
    toc: yes
    latex_engine: xelatex
  word_document:
    toc: yes
  html_document:
    theme: cerulean
    toc: yes
    toc_float: yes
---

### Title: "CUNY SPS MDS DATA607_Tyderverse Create & Extend"
### Author: "Charles Ugiagbe (create) and Mark Schmalfeld (extend)"
### Date: "11/23/2021"


```{r global_options, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<style>
div.aquamarine { background-color:#d9eff9; border-radius: 10px; padding: 5px;}
</style>
<div class = "aquamarine">


## Introduction

Tidyverse is just a collection of R packages underlying same design philosophy, grammar, and data structure. There are currently 8 packages in the `tidyverse` package bundle including:

  * `dplyr`: a set of tools for efficiently manipulating datasets;
  * `forcats`: a package for manipulating categorical variables / factors;
  * `ggplots2`: a classic package for data visualization;
  * `purrr`: another set of tools for manipulating datasets, specially vecters, a complement to `dplyr`;
  * `readr`: a set of faster and more user friendly functions to read data than R default functions;
  * `stringr`: a package for common string operations;
  * `tibble`：a package for reimagining data.frames in a modern way;
  * `tidyr`: a package for reshaping data, a complement to `dplyr`.
  

</div> \hfill\break  
  
**In this assignment, I will use some handy functions in tidyverse package to perform some Analysis**


```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

##       Dataset

The dataset in this project is called “student performance” from https://www.kaggle.com/datasets; The dataset contains a sample of 1000 observations of 8 variables.


I use `read.csv` function to import the csv file to R.

```{r}
url <- "https://raw.githubusercontent.com/omocharly/DATA607_PROJECTS/main/StudentsPerformance.csv"
data <- read.csv(url, header = TRUE)
```


```{r data header}
head(data)
```

Glimpse help us to catch sight of the data to see the data structure.

```{r glimpse data}
glimpse(data)
```

##   dplyr::rename()

rename() changes the names of individual variables using in a column with a new one


```{r rename col}
data1 <- data %>% rename(race = race.ethnicity, parental_Educatn_level= parental.level.of.education, test.prep = test.preparation.course)
head(data1)
```

##   dplyr::select()

Select(): is use for selecting a range of consecutive variables or taking the complement of a set of variables

```{r selection initial example}
data2 <- data1 %>%
  select(gender, math.score, reading.score, writing.score)
head(data2)
```
```{r select examples}
data2b<- data1 %>%
  select( gender,test.prep,math.score,reading.score,writing.score)
head(data2b)
tail(data2b)
```



##   dplyr::filter()

I use the filter() function to filter maths, writing and reading scores that are greater than 95

```{r filter examples}
data3 <- data2 %>%
  filter(math.score == 100, writing.score > 95, reading.score > 95)
data3
```


##    dplyr::arrange()

arrange(): orders the rows of a data frame by the values of selected columns.

```{r arrange examples}
data4 <- data2 %>% arrange(desc(math.score))
head(data4)
```

##    dplyr::mutate() 

mutate() adds new variables that are function of the existing ones to the table and also preserves existing ones.

```{r mutate examples}
data5 <- data4 %>%
  mutate(avg.score = (math.score + writing.score + reading.score) / 3)
head(data5)

data5<- data5 %>%
  mutate(avg.read_write_score =(writing.score+reading.score)/2)
head(data5)

```

## dplyr::case_when() 
Case_when: Function allows you to vectorize multiple if_else() statements. It is an R equivalent of a SQL CASE WHEN statement. 

```{r data mutate}
data6 <- data5 %>% 
  mutate(pass_fail_grade = case_when(avg.score >= 85 ~ 'Pass'
                                     ,TRUE ~ 'Fail' )
         )
head(data6)
```

##    dplyr::summarize() 

```{r ggplot summarize}
data %>% group_by(gender) %>%
  summarize( math_score = sum (math.score)/ n())

data %>% group_by(parental.level.of.education) %>%
  summarize(math_score= sum(math.score)/n())

```

##    Tidyverse::ggplot2()

ggplot2 is a system for 'declaratively' creating graphics, based on "The Grammar of Graphics".

```{r distribution plot}
ggplot(data = data6, aes(x = gender, y = avg.score, col = gender), col = red) + geom_boxplot() + labs(title="Distribution of Students Average score") + theme(plot.title = element_text(hjust=0.5))
```


##### ggplot2 system has many different plot designs that you can create using the appropriate geom_type for the plot.  Below are two simple examples to create a scatter plot evaluation the relationship between reading and math scores by gender (with linear correlation model line using geom_smooth) and a histogram plot of the math scores.

```{r ggplot2 scatter plot with smooth line and histogram}
ggplot(data = data1, aes(x = reading.score, y= math.score, col = gender), col = red) + geom_point() + labs(title="Scatter plot of math score vs reading score by student gender") + theme(plot.title = element_text(hjust=0.5))+geom_smooth(method=lm)

ggplot(data = data1, aes(x = math.score), col = red) + geom_histogram(binwidth = 5) + labs(title="Histogram of math scores") + theme(plot.title = element_text(hjust=0.5))

```



##   map() function use from purrr 
 
Use of the purr Map function to calculate the mean vs a traditional approach.


```{r traditional mean calculation for scores and in this case example is filtered by gender}
data3a<-data2 %>%
  filter(gender== "female",math.score>0, writing.score >95, reading.score >95)
data3a
data3b<-data2 %>%
  filter(gender == "male", math.score>0, writing.score >95, reading.score >95)
data3b


paste0 (" The average female math score is ", (mean(data3a$math.score))) 
paste0 (" The average male math score is ", (mean(data3b$math.score)))


```


```{r use of map() capability to calculate mean and standard deviation}

data2a <- data1 %>%
  select(math.score, reading.score, writing.score)
head(data2a)

paste0 ("The mean math score, reading score and writing score")
map_dbl (data2a, mean)
paste0 ("The stardard deviaiton for math, reading and writing scores")
map_dbl(data2a, sd)


```
##  purrr models short cut to evaluate a correlation

Use of Purr for models and to evaluate the model correlation 

```{r models evaluation with purrr }

models<- data1 %>%
  split(.$gender) %>%
  map(~lm (math.score ~ reading.score, data= .))

models %>%
 map(summary) %>% 
 map_dbl(~.$r.squared)



models2<- data %>%
  split(.$parental.level.of.education) %>%
  map(~lm (math.score ~ reading.score, data= .))

models2 %>%
 map(summary) %>% 
 map_dbl(~.$r.squared)






```

## dyplr  unite() to merge different character columns 

Use of unite() function to merge character columns. 
Merge - gender, level of education and race/ ethnicity.


```{r create a united set of character columns under new column "Merged"}


data<- data %>%
  unite("Merged", parental.level.of.education:gender, remove=FALSE)
head (data)


```




**Other usage of Tidyverse can be found in the textbook "R for Data Science" and other online resource.**
